Efficiency

### 2018-01-08 DraftExec.Rnw
next

The full translation this initial intrigue into the full substance of this paper has come about due to the determination and interest of Gary Johnson.  Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification, and confidence, to meaningfully inform debate.

@@

The same consideration applies to other types of vessels, which might use water-jets.  Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds. However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.  If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt. (Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.

@@

This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.  Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.  There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load. It also corresponds to scenarios where engineers and manufacturers seem from discussion to be far less confident in making quantitative claims about the performance of their own systems.  (They are perfecly happy to make qualitative claims, "it would be alright", "nothing very bad would happen" etc.)

Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load, but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.  Consequently, there are relevantly few pieces of literature which address the present question directly. (Happily, not zero)  However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.

@@

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.  It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal, and their peak output.

The difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is practically unlimited for all intents and purposes on a given voyage. Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.

\subsection{Xenon poisoning and low power limitations}
Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays into Xenon-135, which a half-life of about nine hours.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which cause the continued fission reactions in Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  On the other hand, there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).  This keeps the total level of Xenon in check.

The situation becomes much more complicated when the reactor undergoes a sudden change in power output.  If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.  With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.  Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.  (This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \cite{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.

@@

This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.  For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.  Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.  Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems.    Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful.

@@

\subsection{Recent Commentary}
Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet, but on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

The authors of this piece also suggest that the high cost and weight of the nuclear reactor contributes to the desire to minimise hotel load.  Given the way that the square law drives drag as discussed earlier, driving a cube law for power, there is no chance at all that the hotel load has any bearing on the reactor power required.  The last few knots of top speed would demand ten times over the power required for hotel load, as can be seen in Figure \ref{fig:power_curve}.

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.  One would think that the sorts of emergencies or accidents which would drive this are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably in a couple of days.

This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.  It might be presumed that surfacing and remaining surfaced for a couple of days is likely to end combat operations for a submarine that was located in theatre.  However, remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted, especially since hotel load, in the case of the submarine they propose, amounts to more than propulsion at likely patrol speeds.  As we can gather from other public estimates of submarine submerged endurance \parencite{buckingham2008submarine}, roughly a couple of days is a very significant fraction of the entire dived endurance of a diesel-electric submarine, perhaps 50\%.  It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of diesel-electric submarines, in addition to the reactor.

Consequently, it seems much more likely that the sort of battery-requirements that are necessary for this kind of safety system are not closely linked to a full operational hotel load, but some substantially reduced survival load.  As discussed in later sections, many of the most energy intensive systems in the hotel load that could readily be switched off in an emergency to save power, if saving power was what was required to get through the emergency, including the combat system.  It is difficult to imagine what sort of emergency might require a reactor to be shut down, that wouldn't also justify shutting down essentially every other component of normal hotel load which could plausibly be delayed for a couple of days.

If it was known that the reactor couldn't be restarted after the poison-out period, the mission need to be aborted in any case, and some kind of rescue sought. Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine literally at the entrance of a home port.  It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.

@@

Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.  It says that the total energy in a connected body of fluid is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.

@@

\subsubsection{Venturi Effect}
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is a simple embodiment of the Venturi effect.

\subsubsection{Flow Diffusion}
The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.  This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion \parencite{hamilton1997}.

\subsubsection{Energy waste and efficiency}
The conservation of energy also has useful implications for how the efficiency of systems is thought about.  In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.  A perfectly efficient system won't do any work that isn't for the intended purpose.  In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.  Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.  Noise in the water also reflects energy which is wasted.

@@

Imagine what happens when something moves very fast through water.  The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.  If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.  (This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.  Soon the water catches up, and the bubble implodes with a pop, leaving only tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.  Whilst some energy turns to heat (the remaining steam in the tiny bubbles) quite a bit is propagated away as a sound-wave generated by the implosion.

@@

\subsection{Pressure, Thrust, and Momentum change}
It is worthwhile spending just a moment to touch on the relationship between a few different concepts which are closely related.  Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.  Generating thrust is consequently a key objective of any propulsion system.

Pressure is a measure of force divided by an area over which it is applied. Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.  In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides of an object.  For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.  Being generally backwards facing this acts to push the propeller forwards.  This pressure acts equally to push water backwards, as well as to push the propeller forwards.  Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.  It is the sum of all the pressure differences on the propeller that result in a net force.

Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.  In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.  Submerged jets represent a particular case in point.  A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.  When submerged, the backwards pressue appears dramatically reduced.  In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle. This effect represents one of the important distinctions between the jets which release their water below or above the water-line.

It is also an important law of physics that force or thrust is equal to the rate of change of momentum. Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.  Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust.  Circular movements necessarily are cancelling, and produce no thrust. It is useful to recall however that this principal is simply another complete and way of measuring the consequences of the net pressure differences discussed earlier.  The total rate of change of momentum in the water-column which is induced as a conseqence of the pressure differentials on all sides of the propeller must necessarily be equivalent to the total net force that is produced.  Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.

@@

In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the pressure is increasing in the direction of movement, which means the flow is slowing down as it moves.  At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.  In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.

@@

Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.  Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.  This is the reason that golf balls have dimples, and tennis-balls have fur.

\section{Basics of Ducted Propellers and Pumpjets}
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.

>>>>

The development of the full substance of this paper, has come about due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we shared had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification, and confidence, to meaningfully inform debate.

@@

The same consideration applies to other types of vessels, which might use water-jets.  Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds. However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.  If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt. \footnote{The physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.}  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.

@@

This imposes a constraint on the way in which literature on jets, propellers, and propulsion systems in general is read, since almost all of it is deeply concerned with how a particular system would operate at one particular design speed, and choosing the optimum system for that design speed and load.  Considerable attention is given to how a propeller would work in off-design conditions relating to changes in load, such as in cases when a ship is more lightly laden, or a tug-boat is pushing a different sized ship.  There is, however, a relative scarcity of attention given to the performance of a propulsion system operating at a dramatically different speed, with the same load. Even engineers and manufacturers of propulsion systems tend to be far less confident in making quantitative claims about the performance of their own systems in extreme off-design conditions, since normally decisions affecting system choice are made based on some threshold of performance change which occurs much closer to the nominal design speed or load.

Given that submarines almost by definition (since their hull shape doesn't change, and they must carry ballast to make up for any under-loading) are essentially always operating with the same load (total drag force at any given speed), but have to operate over an extremely wide variety of different speeds (a factor of four or more in variance) with severe efficiency concern at all of them, they represent a truely unique engineering question, seldom discussed in commercial applications.  Consequently, there are relevantly few pieces of literature which address the present question directly. However, there are plenty of instances where the relevant physics and trade-offs are explained in depth to address related, but not identical questions.

@@

In contrast, the Collins Class Submarine's main motor is rated at less than 6MW, with designs for the future submarine appearing to be only slightly larger \parencite{patrick2012}.  It is fair to say that in terms of maximum power ouput, nuclear submarines could have something like 5-10 times as much power at their disposal than diesel-electric submarines would, even at their peak output.

However, the difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage. Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.  This has considerable consequences for the design of all of the submarine systems, including the propulsion system.

\subsection{Xenon poisoning and low power limitations}
Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays quickly into Xenon-135 (which has a half-life of about nine hours), and it is this process which produces the majority of Xenon in the reactor.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which trigger the continued fission reactions of Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).  This keeps the total level of Xenon in check.

The situation changes suddenly when the reactor undergoes a sudden change in power output.  If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.  With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.  Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.  (This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \cite{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.

@@

This effect has a dramatic impact on nuclear submarine design, since much of their design is oriented around being able to comfortably disperse large amounts of excess power, rather than with conserving it.  For the propulsion system, this actually makes having an inefficient propulsor at low speeds a considerable advantage.  Since the reactor will likely need to dispose of excess power, particularly during ramp-down, an inefficient propulsion system actually provides a useful power sink.  Since any excess power will have to be disposed of by some other means (normally by pumping more water to remove the power as the heat) inefficiency at low speeds has no penalty, and probably a marginal benefit, since it will reduce overall demand for additional systems such as pumps to remove the excess heat.  Provided the excess turbulence inside the pumpjet isn't too noisy, wasting energy through the propulsor is useful whenever there is an ample surplus, which could be the case whenever operating noticeably below top speed. Whilst reactors can reach considerably lower power outputs safely in when there are low levels of Iodine and Xenon in the stocks are low (after a smooth ramp-down) the requirement to reduce power gradually means that when speed is reduced suddenly (relative to the 9-hour half life of Xenon 135), for example on arrival in theatre or a higher threat area, this high surplus power will be unavoidable.

@@

\subsection{Recent Commentary}
Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet. However on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

The authors of this piece seem to imply that the relatively high impact of hotel load at patrol speeds is linked to the high cost and weight of the nuclear reactor, and hence this contributes to the desire to minimise hotel load.  Given the way that the square law drives drag as discussed earlier, driving a cube law for power, there is no chance at all that the hotel load has any bearing on the reactor power capacity required.  The last few knots of top speed would demand ten times over the power required for hotel load, as can be seen in Figure \ref{fig:power_curve}.

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.  One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably in a couple of days.

This puts some serious constraint around type of contingency that needs to be planned for in a reactor shutdown event.  It might be presumed that surfacing and remaining surfaced for a couple of days is likely to constitute an end to combat operations for a submarine that was located in theatre, presumeably through enemy action.  However, the implied alternative remaining submerged and continuing combat operations, even at low speeds will essentially necessitate replicating the full power demand of any diesel-electric submarine until such a time as the reactor can be restarted. This fact is emphasised by the DCNS claim that hotel load exceeds propulsive power at patrol speeds.  As we can gather from other public estimates of submarine submerged endurance \parencite{buckingham2008submarine}, roughly a couple of days is a very significant fraction of the entire dived endurance of a diesel-electric submarine, perhaps 50\%.  It is difficult to imagine that nuclear submarines carry nearly 50\% of the battery load of equivalently sized diesel-electric submarines, in addition to the reactor, simply for safety purposes.

Consequently, it seems much more likely that the sort of battery-requirements that are necessary for this kind of safety system are not closely linked to a full operational hotel load, but some substantially reduced survival load.  As discussed in later sections, many of the most energy intensive systems in the hotel load could readily be switched off in an emergency to save power, if saving power was what was required to get through the emergency, including the combat system.  It is difficult to imagine what sort of emergency might require a reactor to be shut down, that wouldn't also justify shutting down essentially every other component of normal hotel load which could plausibly be delayed for a couple of days.

If it was known that the reactor couldn't be restarted after the poison-out period, the mission would need to be aborted in any case, and some kind of rescue sought. Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine was literally at the entrance of a home port.  It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.

Claims that the power management imperatives of nuclear submarines and diesel electric submarines are even remotely similar at low speeds are generally implausible.  Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.  The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.  To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.

@@

Conservation of energy has a particularly relevant embodiment in fluid flows, which is given it's primary expression in Bernoullis equation, as given in Equation \ref{eq:bernoullis}.  It says that the total energy in a connected body of fluid with no additional work being done on it is constant ($C$), though it can change in form between kinetic energy (movement, $\frac{1}{2}\rho v^2$), gravitational potential energy (it being elevated, $\rho gz$, relevant for liquids, less for gas) the heat energy in the fluid (not shown, not relevant for incompbressible fluids), and the pressure of the fluid $p$.

@@

\subsubsection{Venturi Effect}
In the case of submarines and torpedoes the water doesn't undertake a change in height, since intake and nozzle are generally all in line with the central axis of the submarine or torpedo.  As a consequence, the key relationship in Bernoullis equation is the relationship between the liquids velocity, and its pressure.  The consequence is that when water moves through a pipe (duct, or shroud) it's pressure is inversely related to the square of its velocity.  This means that when a liquid is forced to to travel through a narrowing pipe, it's pressure necessarily decreases as the velocity increases.  This is what underlies the Venturi effect.

\subsubsection{Flow Diffusion}
The inverse process is where a pipe increases in volume, and the flow is forced to slow down in order to fill the wider area, and the pressure correspondingly increases.  This is a process called 'diffusion', and is important to achieving high pressure levels in many types of water pumps, including those which will be particularly relevant for pumpjets for watercraft propulsion, particularly elevating static pressure levels in order to prevent cavitation \parencite{hamilton1997, mccormick1963design, bruce1974}.

\subsubsection{Energy waste and efficiency}
The conservation of energy also has useful implications for how the efficiency of systems is thought about.  In particular, because energy is conserved, identifying inefficiencies in a system necessarily involves identifying where energy goes to doing tasks which aren't useful for the intended purpose.  A perfectly efficient system won't do any work that isn't for the intended purpose.  In the case of analysing the efficiency of propulsion systems, the relevant 'work' is almost always related to moving water backwards to produce thrust.  Moving water in directions other than backwards, including random turbulent flows which wind producing heat rather than thust, are two examples of wasted energy.  Noise in the water also reflects energy which is wasted, though the significance of the acoustic signature is not necessarily indicative of the magnitude of energy wasted.

In general terms, energy waste in propulsion systems will result in turbulence (moving water in random chaotic patters, where the energy soon becomes smoothly distributed as heat) and acoustic energy that is propagated away.  Flow separation incolves the creation of larger-scale vortices and eddies, which might be considered as a special case of turbulence, since the length-scale of the unsteady flows is much larger than that of normal turbulence, for instance in a boundary layer.

On important consequences is that whilst increasing creation of noise, flow separation, or turbulence necessarily represent greater amounts of energy wasted, some increase in any given one does not neccessarily indicate the reduction of efficiency of the entire system, as larger amounts of energy may be being saved through reduced wastage through another means, or vice versa. For example, a squeaky join on a car might make a lot of noise, and it will represent some loss of kinetic energy.  However, the brakes can be applied and absorb vastly larger amounts of energy, essentially silently.

Furthermore, efficiency of a system is necessarily given by the a ratio of useful work to total energy expended. Some trends which show increasing levels of energy wasted might be outweighed by larger increases in the overall amount of productive work being achieved.  Pumpjets, which generally tend to require highly turbulent flows, and even more turblent flows at higher speeds, still can become more efficient as speed inceases, as the amount of useful thrust produced increases even faster.  So, despite the wasteful round-and-round movements increasing with speed, the increase in productive front-to-back movement improves even faster, and total efficiency rises.

@@

To try to conceptualise how it occurs in the context of propulsors, one might imagine what happens when something moves very fast through water.  The front side of the object pushes the water forward, but on the back side the water has to push in to fill the space left behind.  If there’s not enough pressure to push the solid water in fast enough, a gap opens up, with just a few gaseous water molecules (steam) inside the cavity.  (This is also described as water boiling at low pressure.)  However the gap doesn’t stay around for long.  Soon the water catches up, and the bubble implodes with a pop, leaving only very tiny bubbles as a result, which you can see in the wake of almost any boat or ship moving at speed.  Whilst some energy turns to heat (the remaining steam in the tiny bubbles) some is propagated away as a sound-wave generated by the implosion.

@@

\subsection{Pressure, Thrust, and Momentum change}
A few simple related concepts make up the foundations for many of the simplest and enduring theoretical models (Momentum Theory, still widely and routinely used to model 'ideal' propeller efficiencies by considering the propeller simple disc \parencite[169]{carlton2007}) and other quite intuitive concepts which migth be more readily grasped in understanding propulsors.

Thrust is a force, measured in Newtons, which is generally used to refer to the forces acting on a vessel or object by its propulsion system to move it forward in the water.  Generating thrust is consequently a key objective of any propulsion system.

Pressure is a measure of force divided by an area over which it is applied. Pressures are necessarily applied to some extent over all of the surfaces of any object, including due to the atmostpheric pressure of air.  In considering the efficiency of propulsion systems for solid bodies, thrust is always the consequence of some \textbf{net} pressure difference on different sides and surfaces of an object.  For instance, on a propeller blade, the rear-facing side of the blade is the side that pushes the water backwards, and hence has a raised pressure on its surface.  Being generally backwards facing this acts to push the propeller forwards.  This pressure acts equally to push water backwards, as well as to push the propeller forwards.  Importantly, the forward-facing side of the propeller will expereince a lowered pressure, which also serves to pull water backwards, and suck the propeller forwards.  It is the sum of all the pressure differences on the propeller that result in a net force.

Importantly in hydrodynamics, it is important to consider the possibility of other forces being generated other surfaces which may have an impact on the movement of the vessel.  In the presence of the hull, water being drawn by the propeller can cause pressure changes on other surface of the vessel which may assist or impede producing a net positive thrust.  Submerged jets represent a particular case in point.  A famous home experiment suggests that one compare the thrust (backwards pressure) experienced by someone holding a garden hose creating a jet of water when it is in the air, as opposed to submerged in a pool of water.  When submerged, the backwards pressue appears dramatically reduced.  In fact, this phenomenon is due to a the water-jet drawing adjacent water with it under-water, the movement of which exerts a negative pressure on the outside of the nozzle, conteracting the pressure exerted on the inside of the nozzle. This effect represents one of the important distinctions between the jets which release their water below or above the water-line, and generally means that underwater jets have much lower jet velocity ratios than those released over water.

It is also an important law of physics that force or thrust is equal to the rate of change of momentum. Put simply, the speed at which one speeds something up, or the increasing amount of something which begins to move, is always equivalent in magnitude to the thrust that is produced.  Since momentum is a vector, this simple law requires that the momentum change has to net out to one particular direction in order to produce a net thrust, which provides the core principal for Momentum Theory as it applies to propeller and jets.  Circular movements necessarily are cancelling, and produce no thrust. It is useful to recall however that this principal is simply another complete, alternative way of measuring the consequences of the net pressure differences discussed earlier.  The total rate of change of momentum in the system which is induced as a conseqence of the pressure differentials on all sides of the propulsor must necessarily be equivalent to the total net force that is produced.  For the garden hoze in the pool or bucket, the jet rapidly entering the liquid in one direction necessitates new flows of liquid in other directions, reducing net moment changed. This new liquid movement will exert pressure reductions on the nozzle surface surface, which reduce the net momentum change affected by the entire system.  These net forces, and net momentum change are necessarily equivalent and equal in effect.  Consequently literature seeking to optimise propulsion might discuss analysis referring to both of these effects or phenomena, and neither are in conflict.

@@

In order for flow separation to occur, a fluid must be moving against an adverse pressure gradient, which simply means that the static pressure (pressure in the rest frame of the fluid) is increasing in the direction of movement, which means the flow is slowing down as it moves, unless new work is being done on it.  At the boundary layer, where the flow velocity is already reduced due to drag forces experienced near the surface, the adverse pressure gradient can be enough to reverse the flow altogether.  In this case, the fluid flows in reverse near the surface and forms a vortex, and the main flow becomes separated form the surface by the votex.

@@

Flow separation tends to occur earlier for laminar boundary layers than for fully turbulent boundary layers.  Consequently, some systems are able to become more efficient  by deliberately inducing turbulent, where the benefit from reduced friction in the laminar boundary layer is outweighed by the cost suction induced by the larger wake field created by earlier flow separation.  This is the reason that golf balls have dimples, and tennis-balls have fur, to enable them to fly further.  The impact of the increased surface area, and more turbulent flow over the surface, is exceeded by the reduced wake field ( trailing area where the flow is totally separted) behind the ball.

\section{Basics of Ducted Propellers and Pumpjets}
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.  Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.
